---
title: "Introduction to dontology"
author: "Package Developer"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to dontology}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

`dontology` is an R package designed for the structured management of pharmaceutical documentation. It uses an ontological approach to treat documents not as flat text files, but as hierarchical trees of entities stored in a relational database (SQLite). 

Key features include:
- **Hierarchical Rendering**: Styled hierarchical headers and Markdown generation.
- **EAV Metadata Model**: Store arbitrary attributes, data frames (tables), and plots within document sections.
- **Audit Trail**: Built-in tracking of `created_by` and `updated_at` for every entity.
- **Interoperability**: Export/Import entire document hierarchies via JSON for Git version control.

## 1. Environment Setup

The system requires an active SQLite connection. All structural constraints and base classes are defined during the schema initialization.

```{r setup}
library(dontology)
library(DBI)
library(RSQLite)

# Create an in-memory database for demonstration
con <- dbConnect(SQLite(), ":memory:")

# Initialize the schema (tables, triggers, and mandatory seeds)
create_schema(con)
```

## 2. Building a Document Hierarchy

Documents are managed via the `DocumentInstance` R6 class, which inherits from the base `Entity` class.

### Manual Construction
You can build a document step-by-step using the `add_child` helper function.

```{r hierarchy}
# 1. Create the root document
doc <- DocumentInstance$new("PROT_2024_001", con, "Clinical Protocol", class_id = "Document")
doc$set_prop("version", "1.0.2")
doc$set_prop("status", "Draft")

# 2. Add high-level sections
sec1 <- add_child(doc, "SEC_1", "Introduction")
sec1$set_prop("content", "This protocol describes the phase II study of drug X.")

# 3. Add nested sub-sections
sub_sec <- add_child(sec1, "SEC_1_1", "Study Objectives")
sub_sec$set_prop("content", "The primary objective is to evaluate safety and efficacy.")

# Persist all changes to the database
doc$save()
```

### Programmatic Construction
For large structures (like CTD or complex validation reports), you can build the hierarchy from a nested list.

```{r build_from_list}
template <- list(
  "SEC_MAT" = list(label = "Materials", children = list(
    "SUB_REAG" = list(label = "Reagents"),
    "SUB_EQUIP" = list(label = "Equipment")
  )),
  "SEC_PROC" = list(label = "Procedure")
)

build_from_list(doc, template)
doc$save()
```

## 3. Managing Content (Tables and Plots)

The package handles complex metadata by serializing data frames to JSON and saving plots to disk while maintaining a reference in the database.

```{r content}
# Navigate to a specific section
results_sec <- add_child(doc, "SEC_RES", "Experimental Results")

# Store a data.frame as a table
stability_data <- data.frame(
  Timepoint = c("Initial", "3 Months", "6 Months"),
  Assay = c(100.1, 99.8, 99.2),
  pH = c(7.0, 7.0, 6.9)
)
results_sec$set_table("table_data", stability_data)

# Store a plot (requires ggplot2)
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
  p <- ggplot(stability_data, aes(x = Timepoint, y = Assay)) + 
       geom_line(group = 1) + geom_point() +
       theme_minimal()
  
  results_sec$set_plot(p)
}

results_sec$set_prop("content", "I am referring to the [Result section](#SEC_RES).")

results_sec$save()
```

## 4. Reusing Standard Clauses

To maintain consistency across multiple documents, `dontology` provides a library of standardized text blocks.

```{r library}
# Initialize the library with predefined pharma clauses
init_standard_library(con)

# Link a 'Cold Storage' clause to our document via a semantic relation
# This avoids text duplication and ensures regulatory compliance
doc$add_relation("contains", "STP_STORAGE_COLD", sort_order = 99)
doc$save()
```

## 5. Validation and Export

The `render()` method generates Markdown. Since we use `normalizePath()` for images, Pandoc will correctly embed plots into the final document regardless of where the export command is called.

```{r render_export}
# 1. Preview the Markdown (without numbers)
cat(doc$render())

# 2. Export to Word in a temporary directory
# We use tempdir() to keep the project clean during vignette building
tmp_docx <- file.path(tempdir(), "Clinical_Protocol.docx")

export_document(doc, format = "word", file_path = tmp_docx)

# Verify the file was created
if (file.exists(tmp_docx)) {
  message("Word document successfully created at: ", tmp_docx)
}
```

## 6. Traceability and Audit Trail

Every `Entity` automatically tracks its creator and the time of the last modification.

```{r audit}
# Query the audit trail directly from the database
audit_log <- dbGetQuery(con, "
  SELECT instance_id, label, created_by, updated_at, updated_by 
  FROM instances 
  WHERE instance_id IN ('PROT_2024_001', 'SEC_1')
")

knitr::kable(audit_log)
```

## 7. Interoperability (JSON Serialization)

The entire document tree can be exported to a single JSON file. This is ideal for sharing templates or using Git for version control of clinical documents.

```{r serialization}
# Export the full hierarchy to a file
export_to_json(doc, "protocol_snapshot.json")

# Import the hierarchy into a fresh database environment
# new_con <- dbConnect(SQLite(), ":memory:")
# create_schema(new_con)
# import_from_json("protocol_snapshot.json", new_con)
```

```{r cleanup, include = FALSE}
# Cleanup logic for the vignette
if (file.exists("protocol_snapshot.json")) unlink("protocol_snapshot.json")
dbDisconnect(con)
```
